A seamless, client-centric programming model for type safe web applications

### 2014-03-01 paper.tex
Drastically cut down on "choosing a compiler"

After the \lstinline!export! call, we enter the domain of client-exclusive code
with the application of \lstinline!runClient!. This function, execute
computations in the \lstinline!Client! monad which is essentially an
IO monad with cooperative multitasking added on top, to mitigate the fact that
Javascript has no native concurrency support. \lstinline!runClient! does not
return, and is the only function with a return type of \lstinline!App Done!,
which ensures that each \lstinline!App! computation contains exactly one client
computation.

@@

\begin{code}
type CallID = Int
type Method = [JSON] -> IO JSON
type AppState = (CallID, [(CallID, Method)])
newtype App a = (StateT AppState IO a)
  deriving (Functor, Applicative, Monad)
\end{code}

@@

In the face of this multitude of options, why did we settle on Haste? We make
the following demands of a candidate source language:

\begin{itemize}
  \item The language must provide a static type system, since one of our
        primary concerns is to reduce defect rates through static typing of
        the client-server communication channel.
  \item The language must be compilable to both Javascript and a format
        suitable for server-side execution as we want our web applications
        to be written as a single, monolithic program.
  \item We also want our compiler to be practical, in the sense that its output
        should not be disproportionately large or slow. This is more important
        for the client part, where we have to contend with slow connections and
        the overhead of running within an interpreter, than for the server part.
\end{itemize}

Our first criterion rules out the dynamically typed Erlang. The second rules
out Fay which unfortunately lacks critical Haskell features, most notably type
classes, and thus has no one to one server-side counterpart compiler.

This still leaves us with four options: Clean, GHCJS, Haste and UHC. However,
the output from GHCJS is prohibitively large - often in the range of several
megabytes for very simple programs - which rules it out for our use case, in
spite of providing very good compatibility with even low-level GHC features.
Similarly, the code produced by the UHC Javascript backend tends to
be relatively slow due to it lacking the more mature optimization capabilities
of GHC, as well as being in the same ballpark as GHCJS for size. Haste, in
contrast, produces code which is generally one to two orders of magnitude
smaller than that produced by GHCJS or UHC.\ \cite{haste}

This leaves us choosing between Clean and Haste for our client compiler. Most
of our abstractions, both the cooperative multitasking and the type-safe
client-server communication itself, are easily expressed as monads, well
supported in Haskell thanks to do-notation and a culture of monadic design
patterns. Add the fact that one of this work's authors is the original author
of Haste, giving us a better position to troubleshoot any code based thereupon,
and Haste becomes the better choice.

@@

One such tradeoff is the decision to make use of the facilities provided by the
web browser - most prominently the Javascript engine's native garbage
collector - as far as possible rather than implementing parts of the GHC
runtime in Javascript. In the same vein, a conscious effort is made to generate
as ``regular-looking'' code as possible. In some rare cases, this means having
to sacrifice compatibility with low-level GHC features. Most notably, Haste
does not support weak references or pre-emptive multitasking, as these would
require modifications to the runtime system and generated code which confound
the structure of the resulting Javascript and have a detrimental impact on code
size and execution speed.

>>>>

After the \lstinline!export! call, we enter the domain of client-exclusive code
with the application of \lstinline!runClient!. This function executes
computations in the \lstinline!Client! monad which is essentially an
IO monad with cooperative multitasking added on top, to mitigate the fact that
Javascript has no native concurrency support. \lstinline!runClient! does not
return, and is the only function with a return type of \lstinline!App Done!,
which ensures that each \lstinline!App! computation contains exactly one client
computation.

@@

\begin{code}
type CallID = Int
type Method = [JSON] -> IO JSON
type AppState = (CallID, [(CallID, Method)])
newtype App a = App (StateT AppState IO a)
  deriving (Functor, Applicative, Monad)
\end{code}

@@

Essentially, our approach is portable to any language or compiler with the
following properties:

\begin{itemize}
  \item The language must provide a static type system, since one of our
        primary concerns is to reduce defect rates through static typing of
        the client-server communication channel.
  \item The language must be compilable to both Javascript and a format
        suitable for server-side execution as we want our web applications
        to be written as a single, monolithic program.
  \item We want the language to provide decent support for a monadic
        programming style, as we our abstractions for cooperative multitasking
        and synchronous client-server communication are neatly expressible in
        this style.
\end{itemize}

While several of the aforementioned compilers fullfil these criteria, the
choice between them becomes almost arbitrary. We chose to base our solution on
Haste as we, by virtue of its authorship, have an intimate knowledge of its
internal workings, strengths and weaknesses. Without doubt, others may see many
reasons to make a different choice.

@@

One such tradeoff is the decision to make use of the facilities provided by the
web browser - most prominently the Javascript engine's native garbage
collector - as far as possible rather than implementing parts of the GHC
runtime in Javascript. In the same vein, a conscious effort is made to generate
as ``regular-looking'' code as possible. This results in large gains in code
compactness and execution speed - in our experiences, compiler Haste programs
are often an order of magnitude smaller and several times faster than an
equivalent program compiled with some other Haskell to Javascript compiler.

In some rare cases, however, this means having to sacrifice compatibility with
low-level GHC features. Most notably, Haste does not support weak references
or pre-emptive multitasking, as these would require modifications to the
runtime system and generated code which confound the structure of the
resulting Javascript and have a detrimental impact on code size and execution
speed.
